Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-04 manu-draft.qmd
all original text added and references cited with cls

>>>>

## Methods

### Study area

Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (Fig 1). Its watershed encompasses another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. It is subtropical, and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural systems within and surrounding Tampa Bay, including direct removal of habitat (including dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.

### Habitats of Tampa Bay

Addressing the full suite of habitats in a watershed is now recognized as critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed. Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.

Habitats generally described as 'subtidal' include hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] are classified as emergent tidal wetlands. For the purposes of this planning effort, supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* As discussed below, uplands are sub-divided into coastal and non-coastal uplands, based on location relative to the 5-foot contour.

### Approach

Recommended targets were created stepwise with geospatial analyses: 1) quantifying habitat status and historical trends; 2) synthesizing historic habitat restoration efforts; and 3) identifying and defining remaining 'opportunities' for restoration that integrated results from the first two analyses. Final development of restoration targets and goals integrated strata and opportunity area information with an analysis of past restoration implementation to use previous completed projects as a guide to a feasible amount of restoration that could be conducted by partners. This approach will ensure that the estuarine-dependent species and faunal guilds throughout the watershed, as defined in the original approach, will continue to be supported.

### Habitat status and trends

For the majority of subtidal, intertidal and supratidal habitats, primary data derived from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD) were utilized. However, to address data gaps for some habitats, results from special studies were integrated with the primary data sources. These included hard bottom [@Kaufman17; @CSA19], dredged holes [@Griffen05; @Raulerson19] and oyster habitat [@OKeife06]*.*

The source data used to estimate the most current coverage of seagrasses, tidal flats, and oysters was the *Seagrass in 2018* geospatial database [@SWFWMD19]. The bi-annual seagrass monitoring program was initiated in 1988 under SWFWMD's Surface Water Improvement and Management program [@Sherwood17; @Tomasko20]*.* SWFWMD has estimated oyster bed coverage as part of this program since 2014.

The source data used to estimate and map trends in development, emergent tidal wetlands, freshwater wetlands, and native upland habitats was the SWFWMD Land Use Land Cover (LULC) series geospatial database [@SWFWMD18]. This comprehensive database classifies the land use and cover types (natural and developed) pursuant to the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often very difficult, it is possible to accurately distinguish forested wetlands from non-forested wetlands. Therefore, for this analysis, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Similarly, within the target- and goal-setting exercise, uplands were combined in one classification. These classifications were reported for each mapping exercise conducted every 2-3 years from the start of the program 1990 through 2017, the most recent year with available data.

To quantify the extent of tidal creeks, GIS data [@JEI16; @JEI20] was clipped to the Florida Department of Environmental Protection stream segments that were classified as estuarine.

### Restoration database

We quantified past restoration efforts in each of the major habitat types to guide decisions on future targets and goals. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the TBEP. Primary data collected included project name, year, description, lead partner, size (area or length), and latitude and longitude. Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and documenting the knowledge of local professionals. Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.

### Creating and combining opportunity layers

Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework. The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070. The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea level rise. The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed. Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web. Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands. These habitats provide important aquifer recharge and wildlife habitat functions.

Opportunity areas, defined here as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets, were also analyzed. The definition and mapping of opportunity areas is necessary to quantify the "restoration potential" for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions. The most appropriate opportunity areas are generally not developed and located on existing public lands or areas identified for acquisition.

The Land Use Land Cover 2017 geospatial database (Table 1) was used as the baseline for cataloguing existing and opportunity areas for intertidal and supratidal habitats in the Tampa Bay watershed. All FLUCCS classification codes were placed into one of three categories. First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography. Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.

Layers for existing public lands and parcels targeted for acquisition were compiled by combining data from the Florida Natural Areas Inventory (Table 1), consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soils distributions can be used to generally represent historical habitat distribution, and can be used to provide generalized restoration guidelines (e.g., tidal wetlands, freshwater wetlands, and native uplands). @Ries14 created a soils suitability analysis for wetland mitigation and restoration sing data from the USDA Web Soil Survey (Table 1) and classified all soils in the Tampa Bay soils into one of three categories (xeric, mesic, and hydric). The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.

A distinction was made between tidal and freshwater wetland restoration potential by intersecting the combined mesic and hydric soils polygons with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were classified as having tidal wetland restoration potential, while mesic or hydric soils occurring above the 5-foot contour were classified as having freshwater wetland restoration potential.

Within the Tampa Bay region, lands adjacent to waters with average salinity values greater than 18 psu are considered most appropriate for higher salinity mangrove/salt barren restoration, while lands adjacent to waters with average salinity values less than 18 psu are considered most appropriate for lower salinity salt marsh (*Juncus* spp.) restoration [@Eleuterius79]. To estimate the relative restoration potential of mangrove/salt barrens and salt marshes, a regional long-term water quality data set was used to create salinity isohalines, which was then binned into two salinity categories: greater and less than an annual mean of 18 psu (Table 1).

All acquired and created GIS layers (Table 1) were converted to 10m x 10m raster data sets and delimited using a previously established watershed boundary. The reservation, restorable, conservation, acquisition, soils, and salinity isohaline layers were then merged to create a comprehensive set of maps and matrices of all possible combinations of opportunity areas. Within an ArcGIS Pro 2.x GIS environment, the SWFWMD 2017 FLUCCS and 2018 submerged data was classified and extracted to the study area into described categories. The dataset was then spatially assigned to existing and proposed restoration areas, reservation areas, soil type, and salinity level.

## Results

### Habitat status and trends

#### Subtidal habitats

While no trend information is available, the best cumulative estimate of natural hard bottom extent in Tampa Bay is 171 ha (Table 2). Oyster bars covered 69 ha (Table 2) in 2018, and a 30% (6 ha) increase reflected since mapping began in 2014 (Table 3a) probably represents improved ground-truthing and photointerpretation of oyster bar signatures from aerial photography. Twelve artificial reefs in Tampa Bay are managed by Hillsborough, Manatee, and Pinellas Counties. Surface area estimates were not available for the Manatee and Pinellas County reefs, but assuming an average size of 4.2 ha, based on the Hillsborough County reefs, the total coverage of artificial reefs in Tampa Bay is estimated to be approximately 67 ha (Table 2).

Total seagrass coverage has increased by 7,027 ha (75%) during the 30-year period of record (Table 3a), and the most current (2018) estimate of total seagrass meadow coverage in Tampa Bay is 16,452 ha. The 2018 coverage of tidal flats and sand other than beaches in Tampa Bay was 6,564 ha (Table 2). A decrease of 4,496 ha (55%) during the 30-year period of record (Table 3a) is associated with the expansion of seagrass to previously non-vegetated bottom area.

#### Intertidal habitats

Between 1990 and 2017, the suite of emergent tidal wetlands (mangroves, salt barrens, and salt marshes) experienced a net gain of 725 ha (10%, Table 3b). The current estimate of the extent of mangrove forests in Tampa Bay is 6,192 ha. Mangrove forest coverage increased by 684 ha (12%). Salt marshes in Tampa Bay cover 1,844 ha, and coverage increased by 30 ha (2%). However, from 1990 to 2017, it is estimated that a net area of 219 ha of salt marshes converted from salt marsh to mangrove habitat (<https://tbep-tech.github.io/landuse/>). The 2017 estimate of the extent of salt barrens in Tampa Bay is 201 ha, and coverage has increased by 14 ha (7%) during the 27-year period of record (Table 3b). Based on GIS data from Janicki Environmental and Mote Marine Laboratory [@JEI16; @JEI20], the extent of tidal creek habitat in the Tampa Bay watershed is approximately 623 km (Table 2). No trend analysis is available.

#### Supratidal habitats

As determined in the 2017 land use/land cover update, the most current estimate of the extent of non-coastal native upland habitats in the Tampa Bay watershed is 56,899 ha, and the extent of "coastal uplands" (defined as below the 5-foot contour) in the Tampa Bay watershed is 1,465 ha (Table 2). Over the 27-year period of record, the suite of native upland habitats has experienced a net loss of 37,051 ha (39%, Table 3b).

The 2017 extent of freshwater wetlands in the Tampa Bay watershed was 88,917 ha (Table 2). Of this total, forested freshwater wetlands comprised 61,565 ha (69%), while non-forested freshwater wetlands comprised 27,351 ha (31%). From 1990-2017, the suite of freshwater wetlands experienced a net gain of 2,444 ha (3%, Table 3b). There has been a 5,335 ha (24%) increase in vegetated non-forested freshwater wetlands since 1990, while forested freshwater wetlands have decreased by 2,891 ha (4%).

### Habitat restoration

A total of 460 projects were documented between 1971 and 2019 (Table 4), addressing the full range of habitat types, including Estuarine (n=228), Freshwater (n=53), Uplands (n=119), and Mixed (n=60). A total of 1,978 ha have been restored, and 12,930 ha and 42.8 km of linear projects were enhanced during the time period. Forty lead partners were documented as responsible for the projects, although some of these lead partners are departments within the same overall agency. Eighty-nine living shoreline projects, seawall enhancements, and oyster reef module installations along shorelines were inventoried, with a linear footprint of 18.2 km.

### Habitat restoration opportunities

In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable. Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area. Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).

The Tampa Bay watershed included a total of 1,260 km^2^ of existing conservation lands, either publicly owned or in conservation easements. However, excluding subtidal areas owned by the State of Florida results in a total of 816 km^2^ of conservation lands, about 13.9% of the watershed area occurring above the MLLW line. (Fig 3). Proposed conservation lands in the Tampa Bay watershed total 1,254 km^2^. The mapped proposed conservation lands generally link eastward and provide wildlife connectivity to the larger-scale Florida Wildlife Corridor (Fig 3).

Reservation lands, from the MLLW line landward to elevation 5-feet (NAVD 88), currently (2017) include 4,622 ha proposed for conservation in the Tampa Bay watershed, with 3,303 ha of native habitats and 1,319 ha of restorable habitats (Fig 1).

Xeric soils were roughly aggregated in an east-west band through the middle of the watershed (Fig 4). Approximately 2,107 km2 was classified as xeric, while 3,760 km2 was classified as mesic/hydric.

Integration of all of the opportunity layers provides a summary of the available restoration for the TBEP habitats of interest (Table 5). The "Native Habitats" columns show the total current extent as well as the portion of the current extent occurring on existing conservation lands and proposed conservation lands, respectively. The "Restorable Habitats" columns show the total restoration opportunity as well as the portion of the total restoration opportunity on existing and proposed conservation lands. The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands (Table 5). However, there are approximately 627 ha of emergent tidal wetland restoration opportunities on existing conservation lands, including 530 ha applicable to higher salinity mangrove forests and salt barrens, and 16 ha applicable to lower salinity salt marsh (e.g., *Juncus roemerianus*) restoration and creation.

The best estimates of total restoration opportunities for urban shorelines and tidal tributaries are provided by the Tampa Bay Living Shoreline Suitability Model (LSSM) prepared by the Florida Fish and Wildlife Conservation Commission [@Boland18]. There is 2,566 km of shoreline in Tampa Bay, and approximately 33% is recommended for living shoreline enhancement.

### Establishment of goals and targets

Recommended targets were based on habitat status and trends, habitat restoration data, identified restoration opportunities, and current and anticipated trends in development, available funding, and regulations [@Hilderbrand05; @Hobbs07].

The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented. However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle. A 30-year planning horizon (2050) is also identified based upon sea level rise projections developed specifically for Tampa Bay [@Burke19]. The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities. Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).

Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels. Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals. Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea level rise scenarios [@Sherwood14; @Geselbracht15].

## Discussion

Because of multiple stressors such as encroaching development and climate change, habitat protection and restoration priorities should be tempered and "reality tested" by what is actually possible today, and what is possible in the future. Many native and potentially restorable habitats are limited, and there will always be restrictions on the financial resources that can be dedicated to public conservation land acquisition and habitat restoration activities.

This replicable method for setting restoration targets and goals provides a systematic attempt to identify habitat protection and restoration targets that are based on what is actually achievable within those limitations. It focuses on existing opportunities for all habitat types, and what is realistically possible in the future, rather than attempting to mimic previous ecological conditions.

### Habitat trends

When viewed as a whole, the most significant and meaningful trends in the TBEP habitats of interest over the periods of record examined include: 1) the 75 percent gain in seagrasses since 1988; 2) the slight gains in emergent tidal wetlands (10%) and freshwater wetlands (2%) since 1990; and 3) the 39% loss in native upland habitats since 1990.

The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space. Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens. This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea level rise [@Comeaux12].

The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4). Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea level rise. Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.

The decrease in native uplands (Table 3b) is the result of continued development in the Tampa Bay watershed, combined with the lack of regulatory protection of native uplands. Attaining the target and goal will require concerted restoration of native upland habitats on existing conservation lands, as well as new conservation lands to offset the continued loss of these habitats to development, and amendments to existing planning, zoning and land development policies or regulations. While federal and state regulations related to listed species management impart some protection to certain rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats, such as pine flatwoods, are left largely unprotected. Unless local governments in the Tampa Bay watershed improve local protections for native uplands, such as strengthening language within comprehensive plans and development ordinances, this trend will likely continue.

### Habitat restoration

Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation. While restoration activities date to 1971, few projects were completed prior to 1990, and from 1990-2010, an annual mean of 68 ha of habitat was restored. Over the past decade (2010-2019), this rate of restoration project completion has increased to over 81 ha/yr. Assuming that funding levels remain in the same range as the past decade, this annual mean can be used to set reasonable limits on restoration potential and targets.

While existing development areas are not considered feasible for major habitat restoration activities at this time, there are many opportunities to enhance and restore habitat functions and improve coastal resilience in urbanized locations. Examples include the construction of living shorelines, placement of submerged habitat modules along developed urban shorelines and seawalls, and creation of backyard habitats. Tidal tributary restoration could also entail improvements including removal of salinity barriers and filling of dredged channel sections with low dissolved oxygen.

Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by TBEP stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds. Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.

There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits. Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18]. Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.

### Rolling easements, mitigation, and restoration consortium

As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea level rise. Where public acquisition is not possible, other conservation mechanisms need to be explored. Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline. A rolling easement is a dynamic mechanism that "rolls" landward as sea levels rise and cause tidal encroachments onto low-lying coastal uplands [@Titus09]. The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses. Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea level rise.

Wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs have historically been conducted independent of watershed-level planning and monitoring processes. This disconnect has contributed to fragmented implementation and inconsistent compliance monitoring of mitigation projects, as well as historically poor documentation of wetland losses and gains in the Tampa Bay watershed. However, if properly focused and comprehensively coordinated, compensatory mitigation activities could significantly contribute to the attainment of wetland habitat restoration goals and targets for the Tampa Bay estuarine system and its contributing watershed.

TBEP will form a public-private partnership to provide the framework for the development of a coordinated approach linking regulatory (compensatory mitigation) and resource management (publicly funded habitat enhancement, restoration, and establishment) programs in the watershed. The creation of the Habitat Management Consortium is expected to provide benefits such as optimizing and improving the cost-effectiveness of habitat protection, restoration and mitigation activities in the watershed.

### Conclusion

The establishment of targets and goals considering climate change, development, land availability, and past achievements, expands the restoration palette to a more comprehensive list of habitats within the system. If successfully implemented, the 2050 goals would total over 4,000 ha of habitat restoration throughout the Tampa Bay watershed. Land acquisition will be an important component of successful completion of different targets and goals, including the salt marsh restoration target for the first ten years. Land acquisition will also provide new opportunities for outdoor access, given that these projects often have a public recreation component.

Our new approach will continue to engage multiple partner agencies, non-governmental organizations, and private citizens in the successful implementation of the restoration plan. Emphases will include recognition of land types particularly vulnerable to climate change or development stressors and these needs will be communicated to restoration partners. Consistent education, targeted funding opportunities, and reporting will also ensure that these newly established targets and goals lead to successful restoration projects, land acquisition, and enhanced ecosystem services.
